Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-25 ms.tex
random2

\newcommand{\Menc}{M_{\rm enc}}
\newcommand{\rhostar}{\rho_*}
\newcommand{\Mstar}{m_{\star}}
\newcommand{\Mseight}{M_{\star,8}}
\newcommand{\Mbh}[1][]{M_{\bullet#1}}
\newcommand{\Mbheight}{M_{\bullet,8}}
\newcommand{\MbheightExp}{\frac{\Mbh}{10^8 \Msun}}

@@

Continuity of the entropy derivative at the stagnation radius demands that the temperature at the stagnation radius is given by (eq.~[\ref{eq:first_law}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p \tilde{v}_{w}^{2}}{2} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K},
\label{eq:Tanalytic}
\end{align}
where $v_{500} \equiv \tilde{v}_{w}/500$ km s$^{-1}$.

Hydrostatic equilibrium can be used to determine the location of the stagnation radius where $v = 0$ (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
\rs=\frac{G \Mbh}{n \vw(\rs)^2}\left(\frac{9-\Gamma}{2} \frac{\Mstar(\rs)}{\Mbh} +\frac{7}{2}\right),
\label{eq:stag_analytic}
\end{align}
where $n$ is the power-law slope of the density profile at $r = r_{\rm s}$.  If the SMBH dominates the gravitational potential at the stagnation radius then $v_{w} \gg \sigma$ and $r_{\rm s} < r_{\rm soi}$ (eq.~[\ref{eq:rsoi}]).  In this case equation (\ref{eq:stag_analytic}) simplifies to
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where the numerical estimate assumes $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their eq.~6).  In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all.  This makes intuitive sense: gas is injected by stars which are gravitationally bound to the black hole, so outflows only occur if the heating rate per gram $\sim \tilde{v}_{w}^{2}$ exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The SMBH accretion rate is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have made use of equation (\ref{eq:stag_simple}), assuming that $v_{w} \gg \sigma$.  The resulting Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) may be compared to the traditional Bondi rate for accretion onto a point source from an external medium of fixed density and temperature density (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_b^2 \rho(r_b) v_{\rm ff}(r_b),
\label{eq:bondi}
\end{align}
where $r_b \equiv GM/c_{s}^{2}$ is the Bondi radius, $v_{\rm ff} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.  Rewriting equation (\ref{eq:mdot_gas}) as $\dot{M}\simeq 4\pi \rs^2 \rho(\rs) v_{\rm ff}(\rs)/3$,
we note a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $\rb$ is replaced by $\rs$.  When the stagnation radius resides interior to the sphere of influence we indeed have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}} = \frac{14}{3}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).

@@

The above analytic results neglect the effects of heat conduction near the stagnation radius, an assumption we now check.  The importance of conductive heating can be estimated by comparing its magnitude to the nominal level of ``wind" heating:
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  10 \phi/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  Because the stagnation radius is usually well inside the break radius, we have approximated the stellar density as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated case, heat conduction is at most an order unity correction for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.

@@

Radiative cooling, if important, qualitative alters key features of the accretion flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm cool} = n^{2}\Lambda \propto T^{\beta-2}$ (see below) is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \in [-0.7,0.5]$.  If manifest, thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Wang+14}).

Radiative cooling can, however, be stabilized by the other sources of cooling, such as the stellar wind source term $\propto -q c_{s}^{2} \propto -qT$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius equals $q \tilde{v}_{w}^{2}/2$ (neglecting conduction and radiative cooling).  Thermal stability at the stagnation radius can thus be assessed by comparing the nominal wind heating rate to the rate of radiative cooling $|\dot{q}_{\rm cool}| =|\Lambda(T) n^2|$,
\begin{align}
\left.\frac{q \vw^2/2}{|\dot{q}_{\rm cool}|}\right|_{\rm r{s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs}) for the gaseous and stellar densities at the stagnation radius, respectively.  Here $n = \rho/\mu m_p$ and we have approximated the cooling function for $T < 2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}$ assuming solar metallicity gas (\citealt{Draine:2011a}, Fig. 34.1).  To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3\rho kT/2\mu m_p)/\dot{q}_{\rm cool}$ to the free-fall time $t_{\rm ff}$ if one uses the equality 
\begin{align}
\rho(\rs)=\frac{3 q(\rs) \tff(\rs)}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations~\eqref{eq:mdot_analytic} and~\eqref{eq:mdot_gas}.

Thermal stability ($0.5 q \vw^2/|\dot{q}_{\rm cool}| > 1$; $t_{\rm cool} > t_{\rm ff}$) at the stagnation radius thus requires a critical minimum heating rate
\begin{align}
v_{w} > v_{\rm crit} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling2}
\end{align}
Our numerical results in $\S\ref{sec:numerical}$ demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.  The inclusion of $\sigma$ in equation (\ref{eq:cooling2}) results because the top two lines apply only for $\tilde{v}_{w} > \sigma$ since their derivation assumed the existence of a stagnation radius given by equation (\ref{eq:rs_simple}).

>>>>

\newcommand{\Menc}{M_{\rm enc}}
\newcommand{\rhostar}{\rho_*}
\newcommand{\Mstar}{M_{\star}}
\newcommand{\Mseight}{M_{\star,8}}
\newcommand{\Mbh}[1][]{M_{\bullet#1}}
\newcommand{\Mbheight}{M_{\bullet,8}}
\newcommand{\MbheightExp}{\frac{\Mbh}{10^8 \Msun}}

@@

Continuity of the entropy derivative at the stagnation radius where $v = 0$ requires that the temperature at the stagnation radius is given by (eq.~[\ref{eq:first_law}])
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p \tilde{v}_{w}^{2}}{2} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K},
\label{eq:Tanalytic}
\end{align}
where $v_{500} \equiv \tilde{v}_{w}/(500$ km s$^{-1}$).  Hydrostatic equilibrium furthermore determines the value of the stagnation radius (Appendix \ref{app:rs}, eq.~\ref{eq:rs2main})
\begin{align}
\rs=\frac{G \Mbh}{n \vw^{2}|_{\rs}}\left(\frac{9-\Gamma}{2} \frac{M_{\star}|_{\rs}}{\Mbh} +\frac{7}{2}\right),
\label{eq:stag_analytic}
\end{align}
where $n$ is the power-law slope of the density profile at $r = r_{\rm s}$.

If the SMBH dominates the gravitational potential at the stagnation radius then $v_{w} \gg \sigma$ and hence the stagnation radius resides well inside the sphere of influence, i.e.~$r_{\rm s} < r_{\rm soi}$ (eq.~[\ref{eq:rsoi}]).  In this case equation (\ref{eq:stag_analytic}) simplifies to
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \tilde{v}_{w}^{2}} \approx 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where the numerical estimate assumes $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their eq.~6).  In the opposite limit of weak heating ($v_{w} \ll \sigma$), the stagnation radius moves to large radii $\gg r_{\rm soi}$ or may not exist at all.  This result makes intuitive sense: gas is supplied by stars which are gravitationally bound to the black hole, so outflows occur only if the specific heating rate $\sim \tilde{v}_{w}^{2}$ exceeds the specific gravitational binding energy $\sim \sigma^{2}$.

@@

The accretion rate onto the SMBH is given by the integrated mass loss rate interior to the stagnation radius (eq.~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
where we have made use of equation (\ref{eq:stag_simple}; assuming that $v_{w} \gg \sigma$) and have divided our results into core ($\Gamma = 0.1$) and cusp ($\Gamma = 0.8$) galaxies.  The resulting Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\frac{\dot{M}}{\dot{M}_{\rm edd}} &\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
where $\dot{M}_{\rm edd} = 2.6M_{\bullet,8}M_{\odot}$ yr$^{-1}$  is the Eddington accretion rate, assuming a radiative efficiency of ten per cent.  Note the sensitive inverse dependence of the accretion rate on the wind heating rate.  
%To reproduce the estimated values $\dot{M}/\dot{M}_{\rm edd} \lesssim 10^{-3}$ in low-luminosity AGN or quiescent galaxies, we thus require $v_{w} \gtrsim 150M_{\bullet,8}^{0.5}$

@@

Our expression for the mass accretion rate (eq.~[\ref{eq:mdot_analytic}]) may be compared to the traditional Bondi rate for accretion onto a point source from an external medium of specified density and temperature (\citealt{Bondi52}):
\begin{align}
  \dot{M}_{\rm B} =4\pi \lambda r_b^2 \rho(r_b) v_{\rm ff}(r_b),
\label{eq:bondi}
\end{align}
where $r_b \equiv GM/c_{s}^{2}$ is the Bondi radius, $v_{\rm ff} = (GM_{\bullet}/r)^{1/2}$ and $\lambda$ is a parameter of order unity.  By rewriting equation (\ref{eq:mdot_gas}) as $\dot{M}\simeq 4\pi \rs^2 \rho(\rs) v_{\rm ff}(\rs)/3$,
one notices a similarity between equation (\ref{eq:mdot_gas}) and the Bondi formula (eq.~[\ref{eq:bondi}]) if $\rb$ is replaced by $\rs$.  When the stagnation radius resides interior to the sphere of influence we indeed have (eq.~[\ref{eq:stag_simple}])
\begin{align}
  \rs\approx\frac{7}{2}\frac{G \Mbh}{\vwO^2} \approx \frac{14}{3}\frac{GM_{\bullet}}{c_{s}^{2}} = \frac{14}{3}r_{\rm B},
  \label{eq:rs_simple}
\end{align}
where the second equality makes use of equation (\ref{eq:Tanalytic}).

@@

Our analytic results above neglect the effects of heat conduction near the stagnation radius, an assumption we now check.  The ratio of the magnitude of conductive heating to the nominal level of ``wind" heating is given by
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T|_{r_{\rm s}}^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}|_{r_{\rm s}} \tilde{v}_{w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  1.0(\phi/0.1)/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where the second equality makes use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}), and we have approximated $\nabla^{2} \sim 1/r_{\rm s}^{2}$, $\nabla \sim 1/r_{\rm s}$.  Because the stagnation radius is usually well inside the break radius, we have approximated the stellar density as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the saturated case, heat conduction is at most an order unity correction for expected values of the saturation parameter $\phi < 0.1$, justifying its neglect in our analysis.

@@

Radiative cooling, if important, qualitative alters key features of the accretion flow.  Initially hydrostatic gas that cools at constant pressure $P \propto nT$ at the rate $\dot{q}_{\rm cool} = n^{2}\Lambda \propto T^{-2.7}$ is thermally unstable, where $\Lambda(T) \propto T^{-\beta}$ with $\beta \sim -0.7$ is the cooling function (see below).  Thermal instability leads to initially higher black hole accretion rates and enhanced star formation which, when coupled to the flow by feed-back processes, can lead to limit cycle behavior inconsistent with our assumption of steady flow (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}; \citealt{Gan+14}).

Radiative cooling can, however, be stabilized by the other sources of cooling, in particular the term $\propto -q c_{s}^{2} \propto -qT$ in the entropy equation (eq.~[\ref{eq:dsdt}]), which at the stagnation radius is balanced by wind heating $\dot{q}_{\rm heat} = q \tilde{v}_{w}^{2}/2$ (neglecting radiative cooling).  Thermal stability at the stagnation radius can thus be assessed by comparing the external (``wind") heating rate to the rate of radiative cooling $|\dot{q}_{\rm cool}| =|\Lambda(T) n^2|$,
\begin{align}
\left.\frac{\dot{q}_{\rm heat}}{|\dot{q}_{\rm cool}|}\right|_{\rm r_{\rm s}} \simeq
  \begin{cases}
   590 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.76}v_{500}^{7.2}  &, \text{core}\\
   210 \eta_{0.02}^{-1} M_{\bullet,8}^{-0.48}v_{500}^{5.8}  &, \text{cusp},     
  \end{cases}
  \label{eq:cooling2}
\end{align}
where we have used equations (\ref{eq:rhors}) and (\ref{eq:rhostarrs}) for the gas and stellar densities at the stagnation radius, respectively.  Here $n = \rho/\mu m_p$ is the particle number density and we have approximated the cooling function for $T < 2\times 10^{7}$ K as $\Lambda(T) = 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}$ assuming solar metallicity gas (\citealt{Draine:2011a}; Fig. 34.1).  To within a constant of order unity, equation (\ref{eq:cooling2}) also equals the ratio of the gas cooling timescale $t_{\rm cool} \equiv (3n kT/2)/\dot{q}_{\rm cool}$ to the free-fall time $t_{\rm ff}$ if one uses the equality 
\begin{align}
\rho(\rs)=\frac{3 q(\rs) \tff(\rs)}{2-\Gamma}.
\label{eq:rhors2}
\end{align}
that results by combining equations~\eqref{eq:mdot_analytic} and~\eqref{eq:mdot_gas}.

Thermal stability ($\dot{q}_{\rm heat} > \dot{q}_{\rm cool}$; $t_{\rm cool} > t_{\rm ff}$) at the stagnation radius thus requires a critical minimum heating rate
\begin{align}
v_{w} > v_{\rm crit} \simeq
  {\rm max}\begin{cases}
   200 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   200 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}\\
\sigma,     
  \end{cases}
  \label{eq:cooling2}
\end{align}
Our numerical results in $\S\ref{sec:numerical}$ will demonstrate that $\tilde{v}_{w} \gtrsim v_{\rm crit} \sim 200-300$ km s$^{-1}$ is generally a condition for stability at all radii, not just at the stagnation radius.  The requirement that $v_{\rm w} \gtrsim \sigma$ in equation (\ref{eq:cooling2}) results because the top two lines are derived under the assumption that the stagnation radius is given by equation (\ref{eq:rs_simple}).  When $v_{\rm w} \lesssim \sigma$ the stagnation radius moves to larger radii than predicted by by equation (\ref{eq:rs_simple}) or may not exist at all, resulting in much higher gas densities and inevitable cooling instability.
